---
title: "Data import"
format: html
---

## Introduction

```{r init, echo=FALSE, warning=FALSE, message=FALSE}
here::i_am('data-import.qmd') 
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, warning = FALSE, message = FALSE) 
library(neonstore)
library(magrittr) 
library(arrow)
library(aws.s3)
```

This script outlines the process to download macroinvertebrate data from a working database of files released by NEON. The full process for updating this database can be found in the [`NEONSizeData` repository](https://size-spectra-boogie.github.io/NEONSizeData/initiate-db.html). This process uses the `neonstore` package to manage data file download and storage for specific NEON data releases that form the underlying derived data products for size analysis. The database of up-to-date files is stored locally on a single computer and the updating of data files is automated. For downstream analysis, the s3 database will serve as the working data source.

We set the specifications of the s3 bucket. This bucket is publicly readable:

```{r set bucket specs, eval = TRUE, echo = TRUE} 
# set the s3 bucket specs  
bucket = "neonsizedata-392202703749-us-east-2-an"
prefix = "neonstore/parquet"
```

We can view the files available on the s3 bucket using the `aws.s3` package to view the contents.

```{r view bucket}

get_bucket(bucket = bucket,
           prefix = prefix,
           max = 3)
  

```

And we can connect to the s3 bucket to perform actions locally by connecting to the remote bucket:

```{r connect remote bucket, eval = TRUE, echo =TRUE}

s3 = arrow::s3_bucket(bucket = paste0(bucket,"/",prefix,"/"))
remote = neon_remote_db(bucket = s3)

```

and then reading in a file from the remote connection.

```{r read fish files, eval = TRUE, echo = TRUE}

neon_remote(
  table = "inv_taxonomyProcessed-basic",
  product = "DP1.20120.001",
  db = remote
  )  %>%
  dplyr::collect() %>% 
  dplyr::glimpse()

```
