---
title: "full-analysis"
format: html
---

<link rel="stylesheet" type="text/css" href="http://cdn.datatables.net/1.10.5/css/jquery.dataTables.min.css">
<script src="http://code.jquery.com/jquery-2.1.2.min.js"></script>
<script src="http://cdn.datatables.net/1.10.5/js/jquery.dataTables.min.js"></script>
<script type="text/javascript">
         $(document).ready(function() {
             $(".dtable").DataTable();
         } );
</script>

## Introduction

```{r init, echo=FALSE, warning=FALSE, message=FALSE}
here::i_am('full-analysis.qmd') 
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, warning = FALSE, message = FALSE)

# Define a custom knitr hook for collapsing output
knitr::knit_hooks$set(fig_collapse=function(before, options, envir) {
  if (before) {
    # Opening HTML tags for the collapsible container
    paste('<details><summary>Click to view figures</summary><div>')
  } else {
    # Closing HTML tags
    paste('</div></details>')
  }
})

source(here::here('code/00_init.R'))

```

```{r load-clean-data, warning=FALSE, message=FALSE, echo=TRUE, eval=TRUE}

macroDWList <- readRDS(file = here('data/derived-data/macroDWList.rds')) 

# How many sampling years does each taxa-site combination have?
site_taxa_nYear_tab = macroDWList %>% 
  map(~.x %>% 
        dplyr::select(collectYear) %>% 
        distinct %>% 
        unlist %>% 
        length()) %>% 
  bind_rows %>% 
  tidyr::pivot_longer(cols= everything(), names_to = 'site-taxa', values_to = 'nYear') 

site_taxa_nYear_tab %>% 
  knitr::kable(., format = 'html', table.attr = "class='dtable'", col.names = c('Site-Taxa', 'nYear'), escape = FALSE)
```

How many site-taxa combinations have only single year of observations?

`r site_taxa_nYear_tab %>% dplyr::filter(nYear == 1) %>% nrow()`

How many have 3+ years? `r site_taxa_nYear_tab %>% dplyr::filter(nYear >= 3) %>% nrow()`

## Fitting max length models

For the site-taxa combinations with 3+ years of observations we will use the `fishmax` package to estimate body size. For the other groups, we will want to devise another approach to get estimates.

```{r subset-3+-years, echo=TRUE, eval=FALSE}
#### This subsets samples to those with 3+ years and creates a list for 
#### modeling. This is a large file and is held only locally currently. 
#### If running this analysis, this will have to be recreated

# find the taxa with only 3+ years
nYear_3 = which(site_taxa_nYear_tab$nYear >= 3)
# select the taxa with only 3+ years
macroDWList_nYear_3 = macroDWList[nYear_3]

# resample the sizes to reduce file size.
# This is not ideal. We want to be able to use all the data if possible.
set.seed(1312)
macroDWList_nYear_3_samps = macroDWList_nYear_3 %>% 
  map(~slice_sample(.x, n = 500, by = collectYear, weight_by = no_m2, replace = TRUE))

macroDWList_nYear_3_list = macroDWList_nYear_3_samps %>% 
  map(~.x %>% 
  named_group_split(collectYear) %>% 
  map(~.x %>%
        select(sizeClass) %>%
        mutate(sizeClass_est = midpoint_resample_vec(sizeClass)) %>% 
        select(sizeClass_est) %>% 
        unlist))

# place the large file in the ignore folder. 
saveRDS(macroDWList_nYear_3_list, here('ignore/macroDWList_nYear_3_list.rds'))
```

To run this on all \~5200 taxa-site observations takes \~10 hours. We could modify the code to estimate body sizes for each taxon in a hierarchical model that uses total information from all sites & years to estimate body sizes in sites where we only have a single years' observation using partial pooling across all samples.

```{r fit-models, echo=TRUE, eval = FALSE}
# load in the large 3+ years list
macroDWList_nYear_3_list = readRDS(here('ignore/macroDWList_nYear_3_list.rds'))
# set the model names for each site-taxa
mod_names = lapply(names(macroDWList_nYear_3_list), FUN = function(a) gsub("^(\\w+)\\s\\/\\s(\\w+)$", "\\2_\\1",a))

# test the named fit_max_model function
 # x = macroDWList_nYear_3_list[[1]]
 # y = unlist(mod_names[1])
 # debugonce(fit_max_model_named)
 # z =  fit_max_model_named(df = x, mod_name = y, rerun = TRUE, overwrite = TRUE, model_types = 'evt')

# set the multisession options for parallel work
future::plan("multisession", workers = 10)

# walk through each taxa and fit multi gvt models in fishmax
# these models are saved to the ignore/models/ folder
# fitting all ~5200 models with the above multisession plan took 
# ~ 10 hours to complete
furrr::future_walk2(
   macroDWList_nYear_3_list,
   mod_names,
   ~fit_max_model_named(df = .x, mod_name = .y, rerun = TRUE, overwrite = FALSE, model_types = c('gvt','gvt_gumbel')),
   .options = furrr_options(
     seed = 1312
     )
   )

```

We can then extract the sizes

```{r extract-max-size, eval = FALSE, echo=TRUE}

modPaths = list.files(here('ignore/models/'), pattern ='.rds', full.names = TRUE)

## test the get_max wide function
# debugonce(get_max_wide)
# x = readRDS(file = modPaths[1])
# get_max_wide(mod_path = modPaths[1])

## run all the models and save to a file for later
modPaths %>% furrr::future_map(get_max_wide) %>% bind_rows() %>% saveRDS(., here('data/max_size.rds'))

```

### Build a pooled model Negative Binomial count model and test

As we mentioned above, we can consider using a hierarchical approach to this procedure. Also, there is a fundamental difference between the estimand in the above Extreme Value Theory (EVT) models and the data we are passing to the function. The EVT models are using just the maximum values observed across the sampling years to estimate the largest size -- so we are only using something like 3-15 data points. Alternatively, we can try to use all the size information to estimate the maximum body size.

I put together a hierarchical model in stan that uses a negative binomial count model to parameterize the entire body size distribution and then uses Posterior Predictive Simulations to estimate the maximum body size observed using a standardized 'effort'. This model uses partial-pooling across sites to estimate the body size distribution, so we can use information across many sites to inform poorly sampled sites.

Below, we test this model against the models in `fishmax`. I deliberately choose a low value for the site with 1 sample to determine the effect of the pooling.

<details>

```{r simulate-multisite data, echo=TRUE}
## set the simulation parameters
simPars = data.frame(site = c('a','b','c','d','e'),
                     mu = c(10,8,7.8, 3, 9.5),
                     sigma = c(1.0,1.2,1.5,0.5,1.8),
                     k = c(4,3,3,1,2),
                     stringsAsFactors = FALSE)
                     
simPars$n = list(c(100,150,200,110),
                           c(100,200,210),
                           c(500,600,700),
                           c(50),
                           c(200,220))

# show histograms of body size distributions for 10,000 individuals
simList = mapply(FUN = truncnorm::rtruncnorm,mean = c(10,8,7.8, 3, 9.5),sd = c(1.0,1.2,1.5,0.5,1.8),
                 MoreArgs = list(n = 1e4,
                                 a = 0))
# plot the distributions
par(mfrow = c(2,3))
apply(simList,2,hist)

## simulate the sampling process and spit out in stan_data format
sim = simulation(simPars = simPars, seed = 1312)

# aggregate to test the simulated observations
# add the max values observed from 10,000 inds for a sanity check

simAgg = do.call(
  data.frame,
  aggregate(
    x ~ site,
    data = sim$observations,
    FUN = function(z) {
      c(
        mean = mean(z),
        sd = sd(z),
        min = min(z),
        max = max(z)
        )
    }
    )
  ) %>% 
  bind_cols(
    data.frame(
    x.mean.1e4 = apply(simList,2,mean),
    x.sd.1e4 = apply(simList,2,sd),
    x.max.1e4 = apply(simList,2,max)
    )
  ) 

simAgg %>% 
  knitr::kable(.)

```

The x.max from the 10,000 individual simulations have a slightly larger max as expected but everything is pretty close, as expected.

</details>

```{r test-hier_nb, warning=FALSE, message=FALSE}

## pull out the stan_data from sim and add the reference sample #
stan_data = sim$stan_data
stan_data$k_ref = 20L
stan_data$mu_upper <- 200
stan_data$sigma_upper <- 100
stan_data$prior_only = 0

# compile stan model
# mod = cmdstan_model(here('code/stan/neg-bin_count_pooled.stan')) #old nb model before stabilizing priors and initialization values to help sampling
mod = cmdstan_model(here('code/stan/hierarchical_nb_max_size_stable.stan'))

init_list= make_init_list_stable(stan_data, chains = 4, seed = 1312)
# fit the model
negbin_fit = mod$sample(
    data = stan_data,
    seed = 1312,
    chains = 4,
    parallel_chains = 4,
    init = init_list,
    iter_warmup = 1000,
    iter_sampling = 1000,
    adapt_delta = 0.95,
    max_treedepth = 12,
    refresh = 0
  )

parameter_summary <- negbin_fit$summary()

# parameter_summary
# sim$site_truth

```

We can check the $L_{max}$ estimates from all the models and compare using one site. Let's check this for one site. I will not choose the site that we purposefully did a deviation.

```{r test-gvt, eval = TRUE, echo = TRUE}

# select just one site
site_a = sim$observations %>% 
  dplyr::filter(site == 'a') %>% 
  named_group_split(event_within_site) %>% 
  map(~dplyr::select(.x, x) %>% unlist)

fit = fit_max_model(site_a, refresh = 0)

get_max(fit, ci = 0.95) %>% 
  bind_rows(
    parameter_summary[which(parameter_summary$variable == 'max_ref_rep[1]'),
                  c('median', 'q5','q95')] %>% 
    dplyr::mutate(model = 'negbin') %>% 
    dplyr::select(model, `max_fit.50%` = median, `max_lwr.2.5%` = q5, `max_upr.97.5%` = q95)
  ) %>% 
  bind_rows(
    data.frame(
      model = 'sim',
      'max_fit.50%' = unlist(simAgg[1, c('x.max')]),
      'max_lwr.2.5%' = NA,
      'max_upr.97.5%' = NA,
      check.names = FALSE)
    ) %>% 
  knitr::kable(.)

```

These simulations and model comparisons show that we do a pretty good job of capturing the signal in this site. We can't fit a model with `fishmax` because we only have a single sample, but let's check on the hierarchical model which can use only a single year.

```{r shrinkage-check}
# select just one site
parameter_summary[
  which(parameter_summary$variable == 'max_ref_rep[4]'),
  c('median', 'q5','q95')
  ] %>% 
  dplyr::mutate(model = 'negbin') %>% 
  dplyr::select(model, `max_fit.50%` = median, `max_lwr.2.5%` = q5, `max_upr.97.5%` = q95) %>% 
  bind_rows(
    data.frame(
      model = 'sim',
      'max_fit.50%' = unlist(simAgg[4, c('x.max')]),
      'max_lwr.2.5%' = NA,
      'max_upr.97.5%' = NA,
      check.names = FALSE)
    ) %>% 
  knitr::kable(.)
```

There is a bit of change, we see a `r (5.19-4.854)/4.854*100`% change from the simulation due to the partial-pooling.

```{r full-NEON-analysis, eval=FALSE, echo=TRUE}
macroTaxaList = readRDS(here('data/derived-data/macroTaxaList.rds'))

# x= macroTaxaList[["CAESP5"]]
# get rough mean sizes quickly to set priors on upper mu bound
means = macroTaxaList %>% 
  map(~.x %>% select(sizeClass) %>% unlist %>%  mean(., na.rm = TRUE)) %>% 
  bind_rows %>% unlist
hist(log10(means))
mean(log(means));mean(means);max(means)
# [1] 1.323187
# [1] 5.177346
# [1] 115
sds = macroTaxaList %>% 
  map(~.x %>% select(sizeClass) %>% unlist %>%  sd(., na.rm = TRUE)) %>% 
  bind_rows %>% unlist
hist(log10(sds))
mean(sds, na.rm = TRUE);max(sds, na.rm = TRUE)
#[1] 2.194067
#[1] 26.71454
# we have some means above 100
# we set mu_upper to 200 and sigma_upper to 100
# convert the taxa sample lists in the correct format for the negative binomial model
macroTaxaList %>% 
  future_map(~make_stanData_taxa(.x),
             .options = furrr_options(seed= 1312)
      ) %>% 
  # this is suprisingly a large file. saved to ignore/ folder.
saveRDS(here('ignore/macroTaxaStanData.rds'))

# make a site index list to map back to model output
macroTaxaList %>% 
  future_map(~make_site_index(.x),
              .options = furrr_options(seed= 1312)) %>% 
  saveRDS(., here('ignore/macroTaxaSiteIndex.rds'))

```

```{r run-negbin-taxa}

# set a future plan for multisession fitting if one isn't already set
if(!inherits(plan(), "multisession")){
  if(parallelly::availableCores() == 24)
    future::plan("multisession", workers = 10, gc = TRUE)
  } else{
    future::plan("multisession", workers = 6, gc = TRUE)
}

# compile stan model
# mod = cmdstan_model(here('code/stan/neg-bin_count_pooled.stan'))
# mod = cmdstan_model(here('code/stan/hierarchical_nb_max_size.stan'))
mod = cmdstan_model(here('code/stan/hierarchical_nb_max_size_stable.stan'))
single_mod = cmdstan_model(here('code/stan/single_nb_max_size_stable.stan'))

macroTaxaStanData = readRDS(here('ignore/macroTaxaStanData.rds')) 
# reorder macroTaxaStanData by most rows to fewest
macroTaxaStanData = macroTaxaStanData[order(sapply(macroTaxaStanData, function(x) x$n_obs), decreasing = TRUE)]
taxaNames = names(macroTaxaStanData)


## based on the upper bounds of the initialization values, we set an upper bound of 4.5. 
## posterior checks will need to be done on this variable to ensure we are not getting a lot 
## of mass in this upper bound and also some prior sensitivity check to ensure we are not 
## massively impacting inferences with this very large intersite-count difference prior.
### run tests with one model to test overwrite
# stan_data = macroTaxaStanData[["CHISP8"]]
# # stan_data$mu_upper <- 200
# # stan_data$sigma_upper <- 100
# y = "CHISP8"
# debugonce(fit_negbin_named)
# fit_negbin_named(stan_data = stan_data, taxaName = y, rerun = TRUE, overwrite = FALSE)
# z = CAESP5_negbin$summary()

# fit the models. This will overwrite existing models if overwrite = TRUE. These are saved in the ignore/ folder
# future_walk2(
#   macroTaxaStanData,
#   taxaNames,
#   ~fit_negbin_named(stan_data = .x, taxaName = .y, rerun = TRUE, overwrite = FALSE),
#   .options = furrr_options(seed = 1312, scheduling = Inf)
#   )
# walk2(
#    macroTaxaStanData,
#    taxaNames,
#    ~fit_negbin_named(stan_data  = .x, taxaName = .y, rerun = TRUE, overwrite = FALSE)
#    )
```






```{r get-nb-max}
# load in the model files lists
nbModFiles = list.files(here('ignore/models/'), '.rds', full.names = TRUE)

<<<<<<< HEAD
# extract taxa names from model lists
taxaNames = nbModFiles %>% 
  map(~sapply(strsplit(sapply(strsplit(.x, "/"),"[",10),"_"), "[", 1)) %>% 
  unlist

# load in site index for each taxa
macroTaxaSiteIndex = readRDS(here('ignore/macroTaxaSiteIndex.rds'))
# set the model index order to each taxa
macroTaxaSiteIndex = macroTaxaSiteIndex[taxaNames]

## read in the taxonomy data to join with index
macroTaxonomy = readRDS(here('data/NEON-macros-tax.rds'))

# read in length-mass regression  
lwCoefsDf = read.csv(here('data/macro_lw_coeffs.csv')) #%>% 
  # add in others that don't have l-m below

# add this to the macroTaxonomy to read in later.
# debugonce(LW_coef)
macroTaxonomy = macroTaxonomy %>% 
  LW_coef(., lw_coef = lwCoefsDf, percent = TRUE)
# read in and clean the site lat-long data to merge with index
siteCoords = read.csv(here('data/site_latlong.csv')) %>% 
  bind_rows(data.frame(
    site = c('BLWA', 'FLNT', 'LIRO','PRLA','SUGG','BARC','PRPO','CRAM','TOOK','TOMB'),
    lat = c(32.54153,31.18542,45.99827,47.15909,29.68778,29.67598,47.12984,46.20967,68.63069,31.85343),
    long = c(-87.79815,-84.43740,-89.70477,-99.11388,-82.01775,-82.00841,-99.25315,-89.47369,-149.61064,-88.15887)
  ))
  
# here is the function to prepare the full data frame with estimated maximum mass. This saves the table so we don't have rerun this every time. Takes 20min (too long) to run sequentially.
if(rerun_data){
  if(!inherits('multisession', plan())){
    future::plan('multisession', workers = 10, gc = TRUE)
  }
future_pmap(
  list(nbModFiles, taxaNames, macroTaxaSiteIndex),
  ~extract_max(filePath = ..1, names = ..2, siteIndex = ..3)
  ) %>%
  bind_rows() %>% 
  dplyr::select(-siteID) %>% 
  dplyr::select(acceptedTaxonID, site = siteName, everything()) %>% 
  dplyr::mutate(site = trimws(as.character(site))) %>% 
  right_join(macroTaxonomy, ., by = 'acceptedTaxonID') %>% 
  left_join(., siteCoords, by = 'site') %>% 
  saveRDS(here('data/derived-data/macroTaxaSiteDf.rds'))
} else{
  macroTaxaSiteDf = readRDS(here('data/derived-data/macroTaxaSiteDf.rds'))
}

# use the est_dw() function to convert length estimates to mass 
macroTaxaSiteDf = macroTaxaSiteDf %>%
  mutate(across(c(mean, median, q5, q95), ~est_dw_simp(.x, a,b,formula_type), .names = "dw_{.col}"))

```

Okay, so now we have all the estimated variables we might want to model, let's take a look 
  
```{r single example}
x %>%   
  dplyr::filter(order == 'Coleoptera') %>% 
  dplyr::filter(grepl('max_ref_rep', variable)) %>% 
  ggplot()+
  geom_point(aes(x = lat, y = log10(dw_median), color = acceptedTaxonID))+
  geom_smooth(aes(x=lat, y = log10(dw_median), group = family), color = 'black', se = FALSE, method = 'lm')+
  scale_color_viridis(discrete = TRUE)+
  theme(legend.position = 'none')
```