---
title: "model-troubleshooting"
format: html
---

## Introduction

<link rel="stylesheet" type="text/css" href="http://cdn.datatables.net/1.10.5/css/jquery.dataTables.min.css">
<script src="http://code.jquery.com/jquery-2.1.2.min.js"></script>
<script src="http://cdn.datatables.net/1.10.5/js/jquery.dataTables.min.js"></script>
<script type="text/javascript">
         $(document).ready(function() {
             $(".dtable").DataTable();
         } );
</script>

```{r init, echo=FALSE, warning=FALSE, message=FALSE}
here::i_am('full-analysis.qmd') 
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, warning = FALSE, message = FALSE)

# Define a custom knitr hook for collapsing output
knitr::knit_hooks$set(fig_collapse=function(before, options, envir) {
  if (before) {
    # Opening HTML tags for the collapsible container
    paste('<details><summary>Click to view figures</summary><div>')
  } else {
    # Closing HTML tags
    paste('</div></details>')
  }
})

source(here::here('code/00_init.R'))

```

This page is for documenting the model checking procedures for the maximum size estimation models. The models were all fit originally using the hierarchical negative binomial model `code/stan/hierarchical_nb_max_size_stable`. With these fits, we identified a number of issues, so we are now working backwards to simpler models. I outline below the steps to keep a record for future.

## Divergent transitions

The original models had \~700, or about 50%, that exhibited divergent transitions after warmup. I should have realized at the outset that we are likely to have many taxa with only a single site, so the hierarchical structure of the model is obviously not appropriate. It turned out that \~400 only have a single site.

```{r id-S-of-one}

macroTaxaStanData = readRDS(here('ignore/macroTaxaStanData.rds')) 

macroTaxaStanData %>% 
  map_lgl(~.x$S == 1) %>% sum

```

This is likely a large part of the issue. Which became apparent with the checks below

### Check models with divergences after sampling

Lots of the models exhibit divergences after warming, most often this is \~1% of transitions, other times if can be up to \~20%. Below we go through a diagnostic procedure to better characterize where these divergences occur and hopefully we can better specify/constrain the model to eliminate this issue.

```{r nb-divergences}
nbFiles = list.files(here('ignore/models/'), '.rds', full.names = TRUE)

nbDivergences = nbFiles %>% 
  map(~check_divergences(filePath = .x))
# what is the distribution of divergent #s
nbDivergences %>% map(~.x %>% pluck('num_diverge') %>% unlist) %>% unlist %>% hist
# what is maximum # of divergences
nbDivergences %>% map(~.x %>% pluck('num_diverge') %>% unlist) %>% unlist %>% max

#what is the 75-percentile # of divergences
nbDivergences %>% map(~.x %>% pluck('num_diverge') %>% unlist) %>% unlist %>% quantile(., 0.75)


# separate out the models with divergences
nbDivergentFiles = nbFiles[nbDivergences %>% map(~pluck(.x, 'diverge')) %>% unlist]
# check with a particularly problematic model
highDiverge = nbDivergences %>% 
  map(~.x %>% pluck('num_diverge') %>% unlist) %>% 
  unlist 

taxaDivergeList = nbDivergentFiles %>% 
  map(~sapply(strsplit(sapply(strsplit(.x,"/"), "[", 10), "_"), "[",1))

divergeStanData = macroTaxaStanData[names(macroTaxaStanData) %in% taxaDivergeList]

# how many of the divergence taxa have only a single site?
divergeStanData %>% 
  map(~.x$S) %>% 
  unlist %>%
  keep(~.x == 1) %>% 
  count()
# all sites with only 1 site show divergences

# how many of the divergence taxa have only 2 samples?
divergeStanData %>% 
  map(~.x$K) %>% 
  unlist %>% 
  keep(~.x == 2) %>% 
  count()
```

To address this issue, we built a simpler single site model as that is the most straightforward way to address this issue. This model is `code/stan/single_nb_max_size_stable.stan`. So with this new approach, I refit just the taxa with observations at only a single site.

```{r refit-S-of-one}
# do a clean load of the stan data to ensure we don't mess anything up
## I am only loading the single model here to ensure we don't accidentally overwrite a model we don't want to 
single_mod = cmdstan_model(here('code/stan/single_nb_max_size_stable.stan'))

macroTaxaStanData = readRDS(here('ignore/macroTaxaStanData.rds')) 
# reorder macroTaxaStanData by most rows to fewest
macroTaxaStanData = macroTaxaStanData[order(sapply(macroTaxaStanData, function(x) x$n_obs), decreasing = TRUE)]
taxaNames = names(macroTaxaStanData)

# first identify the locations of the data with only one site
## convert to logical vector for subsetting stan data and names
macroTaxaSingleVec = macroTaxaStanData %>% 
  map_lgl(~.x$S == 1) %>% unlist 

macroTaxaSingleStanData = macroTaxaStanData[macroTaxaSingleVec]
singleTaxaNames = taxaNames[macroTaxaSingleVec]

# model worked with a single taxa (on only two observations)
# model worked with a single taxa on 22 observations

# ran into an issue that interrupted the future session after many files were written
modFiles = list.files(here('ignore/models/'), '.rds', full.names = TRUE)
fileMeta = file.info(modFiles)
targetTime = as.POSIXct("2026-07-22 17:54:00 CDT")
recentFiles = modFiles[fileMeta$mtime > targetTime]

taxaRecent = singleTaxaNames %>% 
  map_lgl(~grepl(.x, recentFiles, ignore.case = TRUE) %>% sum %>% as.logical)

x = macroTaxaSingleStanData[!taxaRecent]
y = singleTaxaNames[!taxaRecent]
# debugonce(fit_negbin_named)
future::plan("multisession", workers = 10, gc = TRUE)
future_walk2(
   x,
   y,
   ~fit_negbin_named(stan_data  = .x, taxaName = .y, rerun = TRUE, overwrite = TRUE),
  .options = furrr_options(seed = 1312, scheduling = Inf)
   )
# one of the models 3 chains failed need to find this somehow :(
```

A few models still have divergences. I will troubleshoot these but I wanted to work on importing and passing along initial results while I work on divergences.


```{r check-ints}
# make a list of initialization values for checking 
init_lists = macroTaxaStanData %>% 
  map(~make_init_list_stable(
    stan_data = .x,
    chains = 4L,
    seed = 1312)
    )
saveRDS(init_lists, here('ignore/nb_init_lists.rds'))
# code to check initialization values
tau_log_lambda_inits = init_lists %>% 
  future_map(
    ~vapply(.x, function(x) x$tau_log_lambda, numeric(1))
  ) %>% 
  bind_rows(.id = 'taxaName') %>% 
  t() %>% data.frame %>% 
  tibble::rownames_to_column('taxaName') %>% 
  mutate(max = pmax(X1,X2,X3,X4, na.rm = TRUE))
hist(tau_log_lambda_inits$max)
```

```{r diagnostics-bad-models}

fit = readRDS(x)

diagnostics <- fit$diagnostic_summary(
  diagnostics = c(
    "divergences",
    "treedepth",
    "ebfmi"
  ),
  quiet = TRUE
)

diagnostics
# a single chain is the main issue here
sampler_diag <- fit$sampler_diagnostics(
  format = "df"
)

divergence_rate <- mean(
  sampler_diag$divergent__
)#17% of transitions had divergences

parameter_names <- posterior::variables(
  fit$draws()
)

key_parameters <- grep(
  paste0(
    "^(",
    paste(
      c(
        "alpha_log_mu",
        "alpha_log_sigma",
        "tau_log_mu",
        "tau_log_sigma",
        "log_mu_site\\[",
        "log_sigma_site\\[",
        "alpha_log_lambda",
        "tau_log_lambda",
        "z_lambda\\[",
        "log_lambda\\[",
        "log_phi",
        "phi",
        "mu\\[",
        "sigma\\["
      ),
      collapse = "|"
    ),
    ")"
  ),
  parameter_names,
  value = TRUE
)

key_parameters

get_divergence_draws <- function(
    fit,
    variables
) {
  draws <- as.data.frame(
    fit$draws(
      variables = variables,
      format = "df"
    )
  )

  diagnostics <- as.data.frame(
    fit$sampler_diagnostics(
      format = "df"
    )
  )

  draw_key <- paste(
    draws$.chain,
    draws$.iteration,
    sep = ":"
  )

  diagnostic_key <- paste(
    diagnostics$.chain,
    diagnostics$.iteration,
    sep = ":"
  )

  diagnostic_row <- match(
    draw_key,
    diagnostic_key
  )

  if (anyNA(diagnostic_row)) {
    stop(
      "Could not match all posterior draws ",
      "to sampler diagnostics."
    )
  }

  draws$divergent__ <-
    diagnostics$divergent__[diagnostic_row]

  draws$energy__ <-
    diagnostics$energy__[diagnostic_row]

  draws$accept_stat__ <-
    diagnostics$accept_stat__[diagnostic_row]

  draws$treedepth__ <-
    diagnostics$treedepth__[diagnostic_row]

  draws
}

divergence_draws <- get_divergence_draws(
  fit,
  key_parameters
)

divergent_draws_df <- divergence_draws[
  divergence_draws$divergent__ == 1,
]

nondivergent_draws_df <- divergence_draws[
  divergence_draws$divergent__ == 0,
]

screen_divergent_parameters <- function(
    fit,
    variables
) {
  dat <- get_divergence_draws(
    fit = fit,
    variables = variables
  )

  is_divergent <- dat$divergent__ == 1

  if (!any(is_divergent)) {
    return(
      data.frame(
        variable = variables,
        n_divergent = 0L,
        ks_distance = NA_real_,
        median_shift_mad = NA_real_,
        divergent_rank_median = NA_real_,
        divergent_rank_iqr = NA_real_
      )
    )
  }

  results <- lapply(
    variables,
    function(variable) {
      x <- dat[[variable]]

      keep <- is.finite(x)

      x <- x[keep]
      div <- is_divergent[keep]

      x_div <- x[div]
      x_ok <- x[!div]

      if (
        length(x_div) < 2L ||
        length(x_ok) < 2L
      ) {
        return(
          data.frame(
            variable = variable,
            n_divergent = length(x_div),
            ks_distance = NA_real_,
            median_shift_mad = NA_real_,
            divergent_rank_median = NA_real_,
            divergent_rank_iqr = NA_real_
          )
        )
      }

      parameter_scale <- stats::mad(
        x,
        constant = 1.4826
      )

      if (
        !is.finite(parameter_scale) ||
        parameter_scale <= 0
      ) {
        parameter_scale <- stats::sd(x)
      }

      if (
        !is.finite(parameter_scale) ||
        parameter_scale <= 0
      ) {
        parameter_scale <- 1
      }

      ks_result <- suppressWarnings(
        stats::ks.test(
          x_div,
          x_ok,
          exact = FALSE
        )
      )

      scaled_ranks <-
        rank(
          x,
          ties.method = "average"
        ) / (length(x) + 1)

      div_ranks <- scaled_ranks[div]

      data.frame(
        variable = variable,
        n_divergent = length(x_div),

        median_divergent =
          stats::median(x_div),

        median_nondivergent =
          stats::median(x_ok),

        ks_distance =
          unname(ks_result$statistic),

        median_shift_mad =
          abs(
            stats::median(x_div) -
              stats::median(x_ok)
          ) / parameter_scale,

        divergent_rank_median =
          stats::median(div_ranks),

        divergent_rank_iqr =
          stats::IQR(div_ranks)
      )
    }
  )

  results <- do.call(
    rbind,
    results
  )

  results[
    order(
      results$ks_distance,
      decreasing = TRUE,
      na.last = TRUE
    ),
  ]
}

divergence_screen <- screen_divergent_parameters(
  fit,
  key_parameters
)

head(
  divergence_screen,
  20
)

make_bayesplot_nuts <- function(fit) {
  diagnostics <- as.data.frame(
    fit$sampler_diagnostics(
      format = "df"
    )
  )

  diagnostic_variables <- setdiff(
    names(diagnostics),
    c(
      ".chain",
      ".iteration",
      ".draw"
    )
  )

  long_list <- lapply(
    diagnostic_variables,
    function(variable) {
      data.frame(
        Chain = diagnostics$.chain,
        Iteration = diagnostics$.iteration,
        Parameter = variable,
        Value = diagnostics[[variable]]
      )
    }
  )

  do.call(
    rbind,
    long_list
  )
}

nuts_parameters <- make_bayesplot_nuts(
  fit
)

top_parameters <- head(
  divergence_screen$variable,
  6
)

top_draws <- fit$draws(
  variables = top_parameters,
  format = "array"
)

bayesplot::mcmc_pairs(
  top_draws,
  pars = top_parameters,
  np = nuts_parameters,
  condition = bayesplot::pairs_condition(
    nuts = "divergent__"
  ),
  off_diag_args = list(
    alpha = 0.35,
    size = 0.6
  ),
  grid_args = list(
    top = "Divergent versus nondivergent draws"
  )
)

bayesplot::mcmc_parcoord(
  top_draws,
  pars = top_parameters,
  transformations = function(x) {
    scale_value <- stats::mad(
      x,
      constant = 1.4826
    )

    if (
      !is.finite(scale_value) ||
      scale_value <= 0
    ) {
      scale_value <- stats::sd(x)
    }

    if (
      !is.finite(scale_value) ||
      scale_value <= 0
    ) {
      return(x - stats::median(x))
    }

    (
      x - stats::median(x)
    ) / scale_value
  },
  np = nuts_parameters
)

bayesplot::mcmc_trace(
  top_draws,
  pars = top_parameters,
  np = nuts_parameters
)

stan_data = macroTaxaStanData[["DRUCOR"]]
stan_data$mu_upper
init_lists = readRDS(here('ignore/nb_init_lists.rds'))
DRUCOR_initList = init_lists[["DRUCOR"]]

```

### Bad models

Some of the models just have too few sites/samples to adequately and reasonably fit the model. We can prescreen these models using some minimum criteria such as \# of sampling events. Below we explore the data and sampling space that we show divergences.

```{r screen-bad-models}
badTaxa = c(
  "DRUCOR",
)
```

### Check posteriors for some taxa with initializations near boundary

Some of the initial values for the `tau_log_lambda` parameters run up against the set boundaries for this parameters. The value is set at an upper range of `4`. This parameter characterizes the across site expectation in log expected counts. We bounded this to avoid drawing prior draws that lead to `Inf` values during sampling. However, if counts vary greatly among sites, our initial values might exceed this limit leading to rejection of some or all chains. For context, this parameter represents a multiplicative difference in counts, therefore with a limit of `4`, as set, corresponds to a 55x difference in counts for a site that is one-standard deviation. I include a step to check the initialization values for better model specification in the future.

```{r boundary-check}

# some of the initializations and posteriors for tau_log_sigma to ensure mass is not near upper bounds
checkList = c(
  "CAESP5", # this runs at tau_log_sigma 4
  "CHISP8", # 2 chains failed for initialization values > 4
  "RHESP4", # 3 chains rejected for initialization > 4
  "CHISP20", # not sure why this failed, read_cmstan_csv() error
  "NEOSP37", # 7.0% divergences in this model
  "CHASP14", # 7.0% divergences
  "RHYNIG", # 2.0% divergences
  "FLUSP", # 4.0% divergencs
  "EPIPRI", # 8.0% divergences
  "DRUCOR", # 17% divergences
  
)
```

```{r initialization checks}
# code to check initialization values
tau_log_lambda_inits = init_lists %>% 
  future_map(
    ~vapply(.x, function(x) x$tau_log_lambda, numeric(1))
  ) %>% 
  bind_rows(.id = 'taxaName') %>% 
  t() %>% data.frame %>% 
  tibble::rownames_to_column('taxaName') %>% 
  mutate(max = pmax(X1,X2,X3,X4, na.rm = TRUE))
hist(tau_log_lambda_inits$max)
## based on the upper bounds of the initialization values, we set an upper bound of 4.5. 
## posterior checks will need to be done on this variable to ensure we are not getting a lot 
## of mass in this upper bound and also some prior sensitivity check to ensure we are not 
## massively impacting inferences with this very large intersite-count difference prior.
### run tests with one model to test overwrite
# stan_data = macroTaxaStanData[["CHISP8"]]
# # stan_data$mu_upper <- 200
# # stan_data$sigma_upper <- 100
# y = "CHISP8"
# debugonce(fit_negbin_named)
# fit_negbin_named(stan_data = stan_data, taxaName = y, rerun = TRUE, overwrite = FALSE)
# z = CAESP5_negbin$summary()

```
