---
title: "model-troubleshooting"
format: html
---

## Introduction

<link rel="stylesheet" type="text/css" href="http://cdn.datatables.net/1.10.5/css/jquery.dataTables.min.css">
<script src="http://code.jquery.com/jquery-2.1.2.min.js"></script>
<script src="http://cdn.datatables.net/1.10.5/js/jquery.dataTables.min.js"></script>
<script type="text/javascript">
         $(document).ready(function() {
             $(".dtable").DataTable();
         } );
</script>

```{r init, echo=FALSE, warning=FALSE, message=FALSE}
here::i_am('full-analysis.qmd') 
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, warning = FALSE, message = FALSE)

# Define a custom knitr hook for collapsing output
knitr::knit_hooks$set(fig_collapse=function(before, options, envir) {
  if (before) {
    # Opening HTML tags for the collapsible container
    paste('<details><summary>Click to view figures</summary><div>')
  } else {
    # Closing HTML tags
    paste('</div></details>')
  }
})

source(here::here('code/00_init.R'))

if(!inherits(plan(), 'multisession')){
ifelse(parallelly::availableCores() == 24, future::plan('multisession', workers = 10, gc = TRUE), future::plan('multisession', workers = 6, gc = TRUE))
}

```

This page is for documenting the model checking procedures for the maximum size estimation models. The models were all fit originally using the hierarchical negative binomial model `code/stan/hierarchical_nb_max_size_stable`. With these fits, we identified a number of issues, so we are now working backwards to simpler models. I outline below the steps to keep a record for future.

## Divergent transitions

The original models had \~700, or about 50%, that exhibited divergent transitions after warmup. I should have realized at the outset that we are likely to have many taxa with only a single site, so the hierarchical structure of the model is obviously not appropriate. It turned out that \~400 only have a single site.

```{r id-S-of-one}

macroTaxaStanData = readRDS(here('ignore/macroTaxaStanData.rds')) 

macroTaxaStanData %>% 
  map_lgl(~.x$S == 1) %>% sum

```

This is a large part of the issue. Which became apparent with the checks below.

### Check models with divergences after sampling

Lots of the models exhibit divergences after warming, most often this is \~1% of transitions, other times if can be up to \~20%. Below we go through a diagnostic procedure to better characterize where these divergences occur and hopefully we can better specify/constrain the model to eliminate this issue.

```{r nb-divergences}
nbFiles = list.files(here('ignore/models/'), '.rds', full.names = TRUE)
taxaNames = nbFiles %>% 
  map(~sapply(strsplit(sapply(strsplit(.x, "/"),"[",10),"_"), "[", 1)) %>% 
  unlist
nbFiles = setNames(nbFiles, taxaNames)
# check for # of divergences
nbDivergences = nbFiles %>% 
  future_map(~check_divergences(filePath = .x))
# what is the distribution of divergent #s
nbDivergences %>% map(~.x %>% pluck('num_diverge') %>% unlist) %>% unlist %>% hist
# what is maximum # of divergences
nbDivergences %>% map(~.x %>% pluck('num_diverge') %>% unlist) %>% unlist %>% max
which(nbDivergences %>% map(~.x %>% pluck('num_diverge') %>% unlist) %>% unlist == 1116)
#what is the 75-percentile # of divergences
nbDivergences %>% map(~.x %>% pluck('num_diverge') %>% unlist) %>% unlist %>% quantile(., 0.75)


# separate out the models with divergences
nbDivergentFiles = nbFiles[nbDivergences %>% map(~pluck(.x, 'diverge')) %>% unlist]
# check with a particularly problematic model
highDiverge = nbDivergences %>% 
  map(~.x %>% pluck('num_diverge') %>% unlist) %>% 
  unlist 


divergeStanData = macroTaxaStanData[names(macroTaxaStanData) %in% names(nbDivergentFiles )]

# how many of the divergence taxa have only a single site?
divergeStanData %>% 
  map(~.x$S) %>% 
  unlist %>%
  keep(~.x == 1) %>% 
  count()
# all sites with only 1 site show divergences

# how many of the divergence taxa have only 2 samples?
divergeStanData %>% 
  map(~.x$K) %>% 
  unlist %>% 
  keep(~.x == 2) %>% 
  count()

```

To address this issue, we built a simpler single site model as that is the most straightforward way to address this issue. This model is `code/stan/single_nb_max_size_stable.stan`. So with this new approach, I refit just the taxa with observations at only a single site.

```{r refit-S-of-one, eval = FALSE}
# do a clean load of the stan data to ensure we don't mess anything up
## I am only loading the single model here to ensure we don't accidentally overwrite a model we don't want to 
single_mod = cmdstan_model(here('code/stan/single_nb_max_size_stable.stan'))

macroTaxaStanData = readRDS(here('ignore/macroTaxaStanData.rds')) 
# reorder macroTaxaStanData by most rows to fewest
macroTaxaStanData = macroTaxaStanData[order(sapply(macroTaxaStanData, function(x) x$n_obs), decreasing = TRUE)]
taxaNames = names(macroTaxaStanData)

# first identify the locations of the data with only one site
## convert to logical vector for subsetting stan data and names
macroTaxaSingleVec = macroTaxaStanData %>% 
  map_lgl(~.x$S == 1) %>% unlist 

macroTaxaSingleStanData = macroTaxaStanData[macroTaxaSingleVec]
singleTaxaNames = taxaNames[macroTaxaSingleVec]

# model worked with a single taxa (on only two observations)
# model worked with a single taxa on 22 observations

# ran into an issue that interrupted the future session after many files were written
modFiles = list.files(here('ignore/models/'), '.rds', full.names = TRUE)
fileMeta = file.info(modFiles)
targetTime = as.POSIXct("2026-07-22 17:54:00 CDT")
recentFiles = modFiles[fileMeta$mtime > targetTime]

taxaRecent = singleTaxaNames %>% 
  map_lgl(~grepl(.x, recentFiles, ignore.case = TRUE) %>% sum %>% as.logical)

x = macroTaxaSingleStanData[!taxaRecent]
y = singleTaxaNames[!taxaRecent]
# debugonce(fit_negbin_named)
future_walk2(
   x,
   y,
   ~fit_negbin_named(stan_data  = .x, taxaName = .y, rerun = TRUE, overwrite = TRUE),
  .options = furrr_options(seed = 1312, scheduling = Inf)
   )
# one of the models 3 chains failed need to find this somehow :(
```

A few models still have divergences. I will troubleshoot these but I wanted to work on importing and passing along initial results while I work on divergences.

## Max estimates are very high in some cases

Initial exploration of the model results suggest that there are some instances when site-level maximums of some species are very unreasonably high. E.g, 20 mm when the largest observed individual across all sites is 11 mm after we sample. 

So I went about reassessing this model to do two things 1) actually account for the binned structure of the data and 2) build in upper bounds on the truncated normal that are flexible, but also do not put probability in mass ranges that are very unlikely and biologically impossible. This does two important things the original model failed to do. First, the artificial jittering created artificial variability when we can simply account for the bin structure. Second, it 

Here is an example using a well sampled example. The current maximum estimates for this taxa using the lower-truncated Normal are >20mm. But the maximum observed size across >90k individuals is 11mm.

```{r calibaetis-example}

CALSP14 = macroDWList[grepl('CALSP14', names(macroDWList))] %>% do.call('rbind',.)

CALSP14 %>% 
  uncount(weights = no_m2) %>% 
  select(sizeClass) %>% 
  unlist %>% 
  hist(.)

x = macroTaxaStanData[[1]] %>% pluck('x')

```

```{r new-binned-mod}

macroTaxaList = readRDS(here('data/derived-data/macroTaxaList.rds'))

# debugonce(make_binned_nb_stan_data)
x = make_binned_nb_stan_data(
  bin_data = macroTaxaList[[1]],
  event_data = NULL,
  upper_multiplier_max = 1.3,
  k_ref = 20L,
  mu_upper = 400,
  sigma_upper = 200,
  prior_only = 0L
  )

stan_data <- x$stan_data

debugonce(make_init_list_binned_nb)
init_list = make_init_list_binned_nb(
  stan_data = stan_data,
  chains = 4L,
  seed = 1312
)

single_mod = cmdstan_model(here('code/stan/single_nb_max_size_bin.stan'))
stan_data$S <- stan_data$site_id  <- NULL

fit = single_mod$sample(
        data = stan_data,
        seed = 1312,
        chains = 4L,
        parallel_chains = 4L,
        
        init = init_list,
        
        iter_warmup = 1500,
        iter_sampling = 1000,
        adapt_delta = 0.99,
        max_treedepth = 12,
        refresh = 0
)

y = fit$summary()

macroTaxaList[[1]] %>% 
  uncount(weights = no_m2) %>% 
  select(sizeClass) %>% 
  unlist %>% 
  hist

## multisite model test
x = make_retention_grid_stan_data(
  bin_data = macroTaxaList[[1]],
  event_data = NULL,
  morin_link = "logistic",

  prior_only = 0L
  )

stan_data <- x$stan_data

# debugonce(make_init_binned_nb)
init_list = make_retention_grid_init(
  stan_data = stan_data,
  chains = 4L,
  seed = 1312
)

multi_mod_ln_full = cmdstan_model(here('code/stan/hierarchical_ln_max_size_binned.stan'))
multi_mod_ln_slim = cmdstan_model(here('code/stan/hierarchcial_ln_max_size_binned_slim.stan'))


fit = multi_mod_ln_full$sample(
        data = stan_data,
        seed = 1312,
        chains = 4L,
        parallel_chains = 4L,
        
        init = init_list,
        
        iter_warmup = 1500,
        iter_sampling = 1000,
        adapt_delta = 0.99,
        max_treedepth = 12,
        refresh = 0
)

y = fit$summary()

macroTaxaList[[1]] %>% 
  uncount(weights = no_m2) %>% 
  select(sizeClass) %>% 
  unlist %>% 
  hist
```

We have moved from the truncated normal because of poor behavior with the binned data structure and the 

```{r ln-softmax-binned-model}

multi_mod_ln_slim = cmdstan_model(here('code/stan/hierarchical_ln_max_size_binned_slim.stan'))

macroTaxaList %>% 
  future_map(~make_retention_grid_stan_data(
  bin_data = .x,
  event_data = NULL,
  morin_link = "logistic",

  prior_only = 0L
  )) %>% 
  saveRDS(here('ignore/macroTaxaStanDatalnFull.rds'))

macroTaxaStanData = readRDS(here('ignore/macroTaxaStanDatalnFull.rds')) %>% 
  map(~.x$stan_data)

# reorder macroTaxaStanData by most rows to fewest for distributing future work
macroTaxaStanData = macroTaxaStanData[order(sapply(macroTaxaStanData, function(x) (x$S + 1) * x$G * x$J + x$K * x$B), decreasing = TRUE)]
## set the taxa names for naming models
taxaNames = names(macroTaxaStanData)


# debugonce(fit_ln_named)
future_walk2(macroTaxaStanData, taxaNames,
~fit_ln_named(
  stan_data = .x,
  taxaName = .y,
  rerun = TRUE,
  overwrite = FALSE
),
.options = furrr_options(seed =1312, scheduling = Inf)
)

```

## Model diagnostics for the ln softmax model

We can run a full model with all returned quantities (grids, probabilities, etc.), or run a slimmed model and later resample a full diagnostic model from the fits for posterior checks.

```{r full-diagnostics}

# load in the fit of a single 'slim' model
fit = readRDS(here('ignore/models/UVASP_lnbin.rds'))
stan_data = macroTaxaStanData[['UVASP']]

multi_mod_ln_full = cmdstan_model(here('code/stan/hierarchical_ln_max_size_binned.stan'))

gq_fit <- multi_mod_ln_full$generate_quantities(
  fitted_params = fit,
  data = stan_data
)

full_diagnostics = gq_fit$summary()
macroTaxaList[['UVASP']] %>% 
  uncount(weights = no_m2) %>% 
  select(sizeClass) %>% 
  unlist %>% 
  hist
```

The full diagnostics samples a number of other quantities we might be interested, but these are mostly model parameters. Of interest would be the `y_rep*` variables that desribe the expected samples count for each bin. In the above example we only have a single site with a single observation of 11 individuals in sizeclass bin 1mm. The estimated mean size of this taxa is `r full_diagnostics %>% dplyr::filter(grepl('size_mean', variable) %>% select(mean) %>% round(3)` [`r full_diagnostics %>% dplyr::filter(grepl('size_mean', variable) %>% select(q5) %>% round(3)` - `r full_diagnostics %>% dplyr::filter(grepl('size_mean', variable) %>% select(q95) %>% round(3)` 95% Credibility Interval]. 

### Extract model fits

```{r prep-data-extract, echo = TRUE, eval = FALSE}
# load in the model files lists
softmaxModFiles = list.files(here('ignore/models/'), '_lnbin.rds', full.names = TRUE)

# extract taxa names from model lists
taxaNames = softmaxModFiles %>% 
  map(~sapply(strsplit(sapply(strsplit(.x, "/"),"[",10),"_"), "[", 1)) %>% 
  unlist

# load in site index for each taxa
macroTaxaSiteIndex = readRDS(here('ignore/macroTaxaStanDatalnFull.rds')) %>% 
  future_map(~.x %>% pluck('mapping') %>% pluck('site') %>% mutate(site_id = as.character(site_id)))
# set the model index order to each taxa
macroTaxaSiteIndex = macroTaxaSiteIndex[taxaNames]

## read in the taxonomy data to join with index
macroTaxonomy = readRDS(here('data/NEON-macros-tax.rds'))

# read in length-mass regression  
lwCoefsDf = read.csv(here('data/macro_lw_coeffs.csv')) #%>% 
  # add in others that don't have l-m below

# add this to the macroTaxonomy to read in later.
# debugonce(LW_coef)
macroTaxonomy = macroTaxonomy %>% 
  LW_coef(., lw_coef = lwCoefsDf, percent = TRUE)
# read in and clean the site lat-long data to merge with index
siteCoords = read.csv(here('data/site_latlong.csv')) %>% 
  bind_rows(data.frame(
    site = c('BLWA', 'FLNT', 'LIRO','PRLA','SUGG','BARC','PRPO','CRAM','TOOK','TOMB'),
    lat = c(32.54153,31.18542,45.99827,47.15909,29.68778,29.67598,47.12984,46.20967,68.63069,31.85343),
    long = c(-87.79815,-84.43740,-89.70477,-99.11388,-82.01775,-82.00841,-99.25315,-89.47369,-149.61064,-88.15887)
  ))
  
x = softmaxModFiles[2]
y = taxaNames[2]
z = macroTaxaSiteIndex[[2]]
debugonce(extract_max_softmax)
extract_max_softmax(filePath = x, name = y, siteIndex = z) %>% 
  dplyr::select(-site_id)

# here is the function to prepare the full data frame with estimated maximum mass. This saves the table so we don't have rerun this every time. Takes 20min (too long) to run sequentially.
if(rerun_data){
  if(!inherits('multisession', plan())){
    future::plan('multisession', workers = 10, gc = TRUE)
  }
future_pmap(
  list(softmaxModFiles, taxaNames, macroTaxaSiteIndex),
  ~extract_max_softmax(filePath = ..1, name = ..2, siteIndex = ..3)
  ) %>%
  bind_rows() %>% 
  dplyr::select(-site_id) %>% 
  dplyr::select(acceptedTaxonID, site = siteID, everything()) %>% 
  dplyr::mutate(site = trimws(as.character(site))) %>% 
  right_join(macroTaxonomy, ., by = 'acceptedTaxonID') %>% 
  left_join(., siteCoords, by = 'site') %>% 
  saveRDS(here('data/derived-data/macroTaxaSiteDf_softmax.rds'))
} else{
  macroTaxaSiteDf = readRDS(here('data/derived-data/macroTaxaSiteDf.rds'))
}



```

```{r get-softmax-data, echo=TRUE, eval= TRUE}
# use the est_dw() function to convert length estimates to mass 
macroTaxaSiteDf =readRDS(here('data/derived-data/macroTaxaSiteDf_softmax.rds')) %>%
  mutate(across(c(mean, median, sd, q5, q95),
                ~est_dw_simp(.x, a,b,formula_type),
                .names = "dw_{.col}")) %>% 
  ## do some fixes of taxa with missing information
  dplyr::mutate(genus = case_when(scientificName == 'Monophylax mono' ~ 'Monophylax',
                                  scientificName == 'Nanocladius (Plecopterocoluthus) sp.' ~ 'Nanocladius',
                                  .default = genus)) %>% 
  dplyr::mutate(lowestClassification = case_when(
    stringr::str_detect(scientificName,'nae') & is.na(genus) & family %in% 'Chironomidae' ~ 'subfamily',
    stringr::str_detect(scientificName, 'dae', negate = TRUE) & is.na(genus) & family %in% 'Chironomidae' ~ 'tribe',
    is.na(genus) & family %in% 'Chironomidae' ~ 'family',
    is.na(subclass) & is.na(order) & is.na(family) & is.na(genus)~ 'class',
    is.na(order) & is.na(family) & is.na(genus) ~ 'subclass',
    is.na(family) & is.na(genus) ~ 'order',
    is.na(genus) ~ 'family',
    specificEpithet %in% c('sp.','gr.') ~ 'genus',
    .default = 'species')) %>% 
  dplyr::select(scientificName, acceptedTaxonID, class, subclass, order, family, genus, specificEpithet, lowestClassification, site, lat, long, variable, dw_mean, dw_median, dw_sd, dw_q5, dw_q95)
```

## Model diagnostics

```{r check-ints, eval = FALSE}
# make a list of initialization values for checking 
if(rerun_data){
  init_lists = macroTaxaStanData %>% 
  map(~\(x){
    if(x$S == 1){
      make_init_list_single_site(
        stan_data = .x,
        chains = 4L,
        seed = 1312
      )
    } else{
      make_init_list_stable(
        stan_data = .x,
        chains = 4L,
        seed = 1312)
    }
    }
    ) %>%
    saveRDS(here('ignore/nb_init_lists.rds'))
} else{
  init_lists = readRDS(here('ignore/nb_init_lists.rds'))
}
# code to check initialization values
tau_log_lambda_inits = init_lists %>% 
  future_map(
    ~vapply(.x, function(x) x$tau_log_lambda, numeric(1))
  ) %>% 
  bind_rows(.id = 'taxaName') %>% 
  t() %>% data.frame %>% 
  tibble::rownames_to_column('taxaName') %>% 
  mutate(max = pmax(X1,X2,X3,X4, na.rm = TRUE))
hist(tau_log_lambda_inits$max)
```

```{r diagnostics-bad-models}

# fit = readRDS(x)

diagnostics <- fit$diagnostic_summary(
  diagnostics = c(
    "divergences",
    "treedepth",
    "ebfmi"
  ),
  quiet = TRUE
)

diagnostics
# a single chain is the main issue here
sampler_diag <- fit$sampler_diagnostics(
  format = "df"
)

divergence_rate <- mean(
  sampler_diag$divergent__
)#17% of transitions had divergences

parameter_names <- posterior::variables(
  fit$draws()
)

key_parameters <- grep(
  paste0(
    "^(",
    paste(
      c(
        "alpha_log_mu",
        "alpha_log_sigma",
        "tau_log_mu",
        "tau_log_sigma",
        "log_mu_site\\[",
        "log_sigma_site\\[",
        "alpha_log_lambda",
        "tau_log_lambda",
        "z_lambda\\[",
        "log_lambda\\[",
        "log_phi",
        "phi",
        "mu\\[",
        "sigma\\["
      ),
      collapse = "|"
    ),
    ")"
  ),
  parameter_names,
  value = TRUE
)

key_parameters

get_divergence_draws <- function(
    fit,
    variables
) {
  draws <- as.data.frame(
    fit$draws(
      variables = variables,
      format = "df"
    )
  )

  diagnostics <- as.data.frame(
    fit$sampler_diagnostics(
      format = "df"
    )
  )

  draw_key <- paste(
    draws$.chain,
    draws$.iteration,
    sep = ":"
  )

  diagnostic_key <- paste(
    diagnostics$.chain,
    diagnostics$.iteration,
    sep = ":"
  )

  diagnostic_row <- match(
    draw_key,
    diagnostic_key
  )

  if (anyNA(diagnostic_row)) {
    stop(
      "Could not match all posterior draws ",
      "to sampler diagnostics."
    )
  }

  draws$divergent__ <-
    diagnostics$divergent__[diagnostic_row]

  draws$energy__ <-
    diagnostics$energy__[diagnostic_row]

  draws$accept_stat__ <-
    diagnostics$accept_stat__[diagnostic_row]

  draws$treedepth__ <-
    diagnostics$treedepth__[diagnostic_row]

  draws
}

divergence_draws <- get_divergence_draws(
  fit,
  key_parameters
)

divergent_draws_df <- divergence_draws[
  divergence_draws$divergent__ == 1,
]

nondivergent_draws_df <- divergence_draws[
  divergence_draws$divergent__ == 0,
]

screen_divergent_parameters <- function(
    fit,
    variables
) {
  dat <- get_divergence_draws(
    fit = fit,
    variables = variables
  )

  is_divergent <- dat$divergent__ == 1

  if (!any(is_divergent)) {
    return(
      data.frame(
        variable = variables,
        n_divergent = 0L,
        ks_distance = NA_real_,
        median_shift_mad = NA_real_,
        divergent_rank_median = NA_real_,
        divergent_rank_iqr = NA_real_
      )
    )
  }

  results <- lapply(
    variables,
    function(variable) {
      x <- dat[[variable]]

      keep <- is.finite(x)

      x <- x[keep]
      div <- is_divergent[keep]

      x_div <- x[div]
      x_ok <- x[!div]

      if (
        length(x_div) < 2L ||
        length(x_ok) < 2L
      ) {
        return(
          data.frame(
            variable = variable,
            n_divergent = length(x_div),
            ks_distance = NA_real_,
            median_shift_mad = NA_real_,
            divergent_rank_median = NA_real_,
            divergent_rank_iqr = NA_real_
          )
        )
      }

      parameter_scale <- stats::mad(
        x,
        constant = 1.4826
      )

      if (
        !is.finite(parameter_scale) ||
        parameter_scale <= 0
      ) {
        parameter_scale <- stats::sd(x)
      }

      if (
        !is.finite(parameter_scale) ||
        parameter_scale <= 0
      ) {
        parameter_scale <- 1
      }

      ks_result <- suppressWarnings(
        stats::ks.test(
          x_div,
          x_ok,
          exact = FALSE
        )
      )

      scaled_ranks <-
        rank(
          x,
          ties.method = "average"
        ) / (length(x) + 1)

      div_ranks <- scaled_ranks[div]

      data.frame(
        variable = variable,
        n_divergent = length(x_div),

        median_divergent =
          stats::median(x_div),

        median_nondivergent =
          stats::median(x_ok),

        ks_distance =
          unname(ks_result$statistic),

        median_shift_mad =
          abs(
            stats::median(x_div) -
              stats::median(x_ok)
          ) / parameter_scale,

        divergent_rank_median =
          stats::median(div_ranks),

        divergent_rank_iqr =
          stats::IQR(div_ranks)
      )
    }
  )

  results <- do.call(
    rbind,
    results
  )

  results[
    order(
      results$ks_distance,
      decreasing = TRUE,
      na.last = TRUE
    ),
  ]
}

divergence_screen <- screen_divergent_parameters(
  fit,
  key_parameters
)

head(
  divergence_screen,
  20
)

make_bayesplot_nuts <- function(fit) {
  diagnostics <- as.data.frame(
    fit$sampler_diagnostics(
      format = "df"
    )
  )

  diagnostic_variables <- setdiff(
    names(diagnostics),
    c(
      ".chain",
      ".iteration",
      ".draw"
    )
  )

  long_list <- lapply(
    diagnostic_variables,
    function(variable) {
      data.frame(
        Chain = diagnostics$.chain,
        Iteration = diagnostics$.iteration,
        Parameter = variable,
        Value = diagnostics[[variable]]
      )
    }
  )

  do.call(
    rbind,
    long_list
  )
}

nuts_parameters <- make_bayesplot_nuts(
  fit
)

top_parameters <- head(
  divergence_screen$variable,
  6
)

top_draws <- fit$draws(
  variables = top_parameters,
  format = "array"
)

bayesplot::mcmc_pairs(
  top_draws,
  pars = top_parameters,
  np = nuts_parameters,
  # condition = bayesplot::pairs_condition(
    # nuts = "divergent__"
  # ),
  off_diag_args = list(
    alpha = 0.35,
    size = 0.6
  ),
  grid_args = list(
    top = "Divergent versus nondivergent draws"
  )
)

bayesplot::mcmc_pairs(
  fit$draws(variables = c('tau_log_mu','tau_log_sigma'),
            format = 'array'),
  pars = c('tau_log_mu','tau_log_sigma'),
  np = nuts_parameters,
  off_diag_args = list(
    alpha = 0.35,
    size = 0.6
  ),
  grid_args = list(
    top = "Divergent versus nondivergent draws"
  )
)

bayesplot::mcmc_parcoord(
  top_draws,
  pars = top_parameters,
  transformations = function(x) {
    scale_value <- stats::mad(
      x,
      constant = 1.4826
    )

    if (
      !is.finite(scale_value) ||
      scale_value <= 0
    ) {
      scale_value <- stats::sd(x)
    }

    if (
      !is.finite(scale_value) ||
      scale_value <= 0
    ) {
      return(x - stats::median(x))
    }

    (
      x - stats::median(x)
    ) / scale_value
  },
  np = nuts_parameters
)

bayesplot::mcmc_trace(
  top_draws,
  pars = top_parameters,
  np = nuts_parameters
)

stan_data = macroTaxaStanData[["DRUCOR"]]
stan_data$mu_upper
init_lists = readRDS(here('ignore/nb_init_lists.rds'))
DRUCOR_initList = init_lists[["DRUCOR"]]

```

### Bad models

Some of the models just have too few sites/samples to adequately and reasonably fit the model. We can prescreen these models using some minimum criteria such as \# of sampling events. Below we explore the data and sampling space that we show divergences.

```{r screen-bad-models, eval = FALSE}
badTaxa = c(
  "DRUCOR",
)
```

### Check posteriors for some taxa with initializations near boundary

Some of the initial values for the `tau_log_lambda` parameters run up against the set boundaries for this parameters. The value is set at an upper range of `4`. This parameter characterizes the across site expectation in log expected counts. We bounded this to avoid drawing prior draws that lead to `Inf` values during sampling. However, if counts vary greatly among sites, our initial values might exceed this limit leading to rejection of some or all chains. For context, this parameter represents a multiplicative difference in counts, therefore with a limit of `4`, as set, corresponds to a 55x difference in counts for a site that is one-standard deviation. I include a step to check the initialization values for better model specification in the future.

```{r boundary-check, eval = FALSE}

# some of the initializations and posteriors for tau_log_sigma to ensure mass is not near upper bounds
checkList = c(
  "CAESP5", # this runs at tau_log_sigma 4
  "CHISP8", # 2 chains failed for initialization values > 4
  "RHESP4", # 3 chains rejected for initialization > 4
  "CHISP20", # not sure why this failed, read_cmstan_csv() error
  "NEOSP37", # 7.0% divergences in this model
  "CHASP14", # 7.0% divergences
  "RHYNIG", # 2.0% divergences
  "FLUSP", # 4.0% divergencs
  "EPIPRI", # 8.0% divergences
  "DRUCOR", # 17% divergences
  
)
```

```{r initialization checks, eval = FALSE}
# code to check initialization values
tau_log_lambda_inits = init_lists %>% 
  future_map(
    ~vapply(.x, function(x) x$tau_log_lambda, numeric(1))
  ) %>% 
  bind_rows(.id = 'taxaName') %>% 
  t() %>% data.frame %>% 
  tibble::rownames_to_column('taxaName') %>% 
  mutate(max = pmax(X1,X2,X3,X4, na.rm = TRUE))
hist(tau_log_lambda_inits$max)
## based on the upper bounds of the initialization values, we set an upper bound of 4.5. 
## posterior checks will need to be done on this variable to ensure we are not getting a lot 
## of mass in this upper bound and also some prior sensitivity check to ensure we are not 
## massively impacting inferences with this very large intersite-count difference prior.
### run tests with one model to test overwrite
# stan_data = macroTaxaStanData[["CHISP8"]]
# # stan_data$mu_upper <- 200
# # stan_data$sigma_upper <- 100
# y = "CHISP8"
# debugonce(fit_negbin_named)
# fit_negbin_named(stan_data = stan_data, taxaName = y, rerun = TRUE, overwrite = FALSE)
# z = CAESP5_negbin$summary()

```
