---
title: "Quick example"
format: html
---

## Introduction

```{r init, echo=FALSE, warning=FALSE, message=FALSE}
here::i_am('quick-ex.qmd') 
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, warning = FALSE, message = FALSE)
library(here)
library(junkR)
library(dplyr)
library(purrr)
library(fishmax)

```

```{r load-clean-data, warning=FALSE, message=FALSE, echo=TRUE, eval=TRUE}

macroDW <- readRDS(file = here("data/macro_dw_raw.rds"))

# clean macros data.frame and aggregate abundances within sites by taxonID
macroDWList = macroDW %>% 
  dplyr::select(siteID, collectDate, sampleID, acceptedTaxonID,sizeClass, dw, no_m2) %>% 
  dplyr::mutate(collectDate = as.Date(collectDate)) %>% 
  dplyr::mutate(collectMonth = as.Date(
    paste0(
      lubridate::year(collectDate),
      "-",
      lubridate::month(collectDate),
      "-01"),
    format = "%Y-%m-%d"),
    collectYear = lubridate::year(collectDate)) %>% 
  dplyr::summarise(no_m2 = sum(no_m2), .by = c(siteID, collectYear, acceptedTaxonID, sizeClass, dw)) %>% 
  named_group_split(siteID, acceptedTaxonID)

```

How many groups have only 1 size class?

`r macroDWList %>% keep(~nrow(.) == 1) %>% length()`

### Example using the taxa with the most observations

Here we walk through a procedure to estimate the maximum body size for each taxa using observations across years. For each taxa, we group all observations within a year to best capture the full potential sizes observed. Below is a histogram of all observed body sizes within this example.

```{r well-sampled-ex, eval = TRUE}
## Which has the most observations?
max_r = map_int(macroDWList, ~nrow(.x)) %>% max();max_r

## practice workflow on a single, well-sampled taxa
x = Filter(function(x) nrow(x) == max_r, macroDWList) %>% purrr::flatten_df()

x %>% 
  ggplot()+
  geom_histogram(aes(x = sizeClass, y = ..density.., weight = no_m2))
```

We then go through a procedure to resample from the size distribution within each year. In this case, the sizes are resampled with replacement to a fixed length (e.g., 1000). We then add variability around the size bin by sampling from a uniform distribution $\pm$ 0.5mm around the length class.

```{r sizes-resamp, eval = FALSE}
x_samp = x %>% 
  slice_sample(n = 100, by = collectYear, weight_by = no_m2, replace = TRUE)

x_list = x_samp %>% 
  named_group_split(collectYear) %>% 
  map(~.x %>%
        select(sizeClass) %>%
        mutate(sizeClass_est = midpoint_resample_vec(sizeClass)) %>% 
        select(sizeClass_est) %>% 
        unlist)

```

From here we fit a suite of models using the size vectors of body sizes. These models are outlined in Heather et al. 2026 ["Beyond the single biggest fish: robust inference of species maximum length ($L_{max}$) from sample maxima"](https://doi.org/10.1007/s11160-026-10047-2). That estimate maximum length using Bayesian inference with three different approaches. These have been collected into an R package `fishmax`. 

```{r est-max-length, eval = FALSE}

fishmax::fit_max_model(x_list)
```

```{r load-mod, eval = TRUE, echo = FALSE}
fit = readRDS(here('data/models/CHISP8_PRLA.rds'))
```

From these models we can look at the estimated maximum lengths with `get_max()`

```{r look-max-ests, eval=TRUE, echo=FALSE}
get_max(fit) %>% knitr::kable(.)
```

and lastly plot the estimates and their uncertainty with `plot_max()`

```{r plot-max-ests, eval = TRUE, echo=FALSE}

plot_max(fit)
```

